---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
activity_data = read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
```{r}
# Use tapply to calculate the total number of steps taken per day
total_steps_per_day = tapply(activity_data$steps, list(activity_data$date), sum, na.rm = TRUE)
total_steps_per_day
```

```{r}
# Use hist() to make a histogram of total steps taken each day
hist(total_steps_per_day, main ="Histogram of total steps taken each day")
```

```{r}
# Calculate mean and median of total_steps_taken_per_day
mean(total_steps_per_day)
median(total_steps_per_day)
```


## What is the average daily activity pattern?
```{r}
# Use tapply to calculate the average (mean) number of steps taken per interval averaged across all days
average_steps_per_interval = tapply(activity_data$steps, list(activity_data$interval), mean, na.rm = TRUE)
# Create a time series plot of the 5-min interval (x-axis) and average number of steps per interval (y-axis)
plot(activity_data$interval[0:length(average_steps_per_interval)], average_steps_per_interval, type = "l", xlab = "interval")
```

```{r}
activity_data$interval[which.max(average_steps_per_interval)]
```

## Inputing missing values

```{r}
# Calculate and report the total number of rows with NAs (i.e. count of which rows are not complete cases)
length(which(!complete.cases(activity_data)))
```

#### The strategy for inputting missing data will be to replace NAs with the corresponding interval averages

```{r}
# Create a new data set equal to the original data set with missing data filled in (use interval averages to replace nas)
new_activity_data = activity_data
new_activity_data$steps[!complete.cases(new_activity_data$steps)] = average_steps_per_interval[!complete.cases(new_activity_data$steps) %% 288]
```

```{r}
# Use tapply to calculate the new total number of steps taken per day
new_total_steps_per_day = tapply(new_activity_data$steps, list(new_activity_data$date), sum, na.rm = TRUE)

# Use hist() to make a histogram of new total steps taken each day
hist(new_total_steps_per_day, main ="Histogram of total steps taken each day")

# Calculate mean and median of new_total_steps_taken_per_day
mean(new_total_steps_per_day)
median(new_total_steps_per_day)
```

##### These values differ from the estimates from the first part of the assignment. Inputting missing data increased the the estimates of the total daily number of steps on average.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
# Create a new factor variable in the dataset with two levels - "weekday"" and "weekend"" indicating whether a day is a weekday or weekend day
new_factor = weekdays(as.Date(new_activity_data$date))
new_factor[which(new_factor == "Saturday" | new_factor == "Sunday")] = "weekend"
new_factor[which(new_factor != "weekend")] = "weekday"
new_activity_data = cbind(new_activity_data, new_factor)
```

```{r}
# Make a panel plot for average steps taken for weekdays or weekends
weekdays_data = new_activity_data[which(new_activity_data$new_factor == "weekday"),]
average_steps_per_weekday = tapply(weekdays_data$steps, list(weekdays_data$interval), mean, na.rm = "TRUE")

weekends_data = new_activity_data[which(new_activity_data$new_factor == "weekend"),]
average_steps_per_weekend = tapply(weekends_data$steps, list(weekends_data$interval), mean, na.rm = "TRUE")

par(mfrow=c(2,1))
plot(new_activity_data$interval[0:length(average_steps_per_weekday)], average_steps_per_weekday, type = "l", xlab = "interval")
plot(new_activity_data$interval[0:length(average_steps_per_weekend)], average_steps_per_weekend, type = "l", xlab = "interval")
```